---
title: "Environment Time-Series Check"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

```{r setup}
library(DBI)
library(RSQLite)
library(dplyr)
library(ggplot2)
library(tidyr)

report_dir <- dirname(normalizePath(knitr::current_input()))
db_path <- file.path(report_dir, "temperatures.db")
environment_table_path <- file.path(report_dir, "environment_table.RDS")

stopifnot(file.exists(db_path))
stopifnot(file.exists(environment_table_path))

conn <- dbConnect(SQLite(), db_path)
temperatures <- dbReadTable(conn, "temperatures")
dbDisconnect(conn)

environment_table <- readRDS(environment_table_path) |>
  select(
    env_series_id,
    target_gamma = one_over_f_gamma,
    temperature_mean,
    temperature_sd,
    environment_replicate
  )

temperatures <- temperatures |>
  left_join(environment_table, by = "env_series_id")

target_temperature_mean <- unique(environment_table$temperature_mean)
target_temperature_sd <- unique(environment_table$temperature_sd)
```

## Overview

```{r overview-table}
overview <- temperatures |>
  summarise(
    rows = n(),
    environments = n_distinct(env_series_id),
    first_time = min(time),
    last_time = max(time),
    target_mean = first(temperature_mean),
    target_sd = first(temperature_sd)
  )

knitr::kable(overview)
```

## Temperature-Series Moments

Each generated time series should have mean close to `temperature_mean` and
sample standard deviation close to `temperature_sd`.

```{r moment-table}
moment_summary <- temperatures |>
  group_by(env_series_id, target_gamma, environment_replicate) |>
  summarise(
    n = n(),
    mean_temperature = mean(temperature),
    sd_temperature = sd(temperature),
    .groups = "drop"
  ) |>
  mutate(
    mean_error = mean_temperature - target_temperature_mean,
    sd_error = sd_temperature - target_temperature_sd
  )

knitr::kable(
  moment_summary |>
    arrange(target_gamma, environment_replicate) |>
    mutate(
      mean_temperature = round(mean_temperature, 6),
      sd_temperature = round(sd_temperature, 6),
      mean_error = signif(mean_error, 3),
      sd_error = signif(sd_error, 3)
    )
)
```

```{r moment-plot, fig.width=8, fig.height=4}
moment_summary |>
  pivot_longer(
    cols = c(mean_temperature, sd_temperature),
    names_to = "statistic",
    values_to = "value"
  ) |>
  ggplot(aes(x = target_gamma, y = value, group = statistic, colour = statistic)) +
  geom_hline(yintercept = target_temperature_mean, linetype = 2) +
  geom_hline(yintercept = target_temperature_sd, linetype = 3) +
  geom_point(alpha = 0.8) +
  facet_wrap(~ statistic, scales = "free_y") +
  labs(
    x = "Specified one_over_f_gamma",
    y = "Realized value",
    colour = NULL
  ) +
  theme_minimal()
```

## Representative Time Series

The plot below shows one environment replicate per value of
`one_over_f_gamma`. To keep the plot readable, it shows evenly spaced sampled
time points rather than every observation.

```{r sampled-time-series, fig.width=10, fig.height=8}
sample_every <- max(1L, floor(max(temperatures$time) / 1000))

representative_series <- temperatures |>
  group_by(target_gamma) |>
  filter(environment_replicate == min(environment_replicate)) |>
  ungroup() |>
  filter((time - min(time)) %% sample_every == 0)

ggplot(representative_series, aes(x = time, y = temperature)) +
  geom_line(linewidth = 0.25) +
  facet_wrap(~ target_gamma, ncol = 2) +
  labs(
    x = "Time",
    y = "Temperature"
  ) +
  theme_minimal()
```

## Spectral Exponents

The generator uses `one_over_f_gamma` as an amplitude-spectrum exponent:
amplitudes scale approximately as `1 / frequency^gamma`. Therefore the power
spectrum scales approximately as `1 / frequency^(2 * gamma)`.

The table below estimates the slope of the log power spectrum for each
environment. `estimated_power_exponent` should be close to
`2 * one_over_f_gamma`; `estimated_amplitude_gamma` is half of that value and
should be close to `one_over_f_gamma`.

```{r spectral-functions}
estimate_spectral_exponent <- function(temperature) {
  y <- temperature - mean(temperature)
  n <- length(y)
  fft_values <- stats::fft(y)
  frequencies <- seq_len(floor(n / 2)) / n
  power <- Mod(fft_values[seq_along(frequencies) + 1L])^2 / n

  keep <- is.finite(power) & power > 0 & frequencies > 0
  fit <- stats::lm(log10(power[keep]) ~ log10(frequencies[keep]))
  power_exponent <- -unname(stats::coef(fit)[[2]])

  tibble(
    estimated_power_exponent = power_exponent,
    estimated_amplitude_gamma = power_exponent / 2,
    spectral_r_squared = summary(fit)$r.squared
  )
}

spectral_summary <- temperatures |>
  group_by(env_series_id, target_gamma, environment_replicate) |>
  summarise(
    estimate_spectral_exponent(temperature),
    .groups = "drop"
  ) |>
  mutate(
    expected_power_exponent = 2 * target_gamma,
    power_exponent_error = estimated_power_exponent - expected_power_exponent,
    amplitude_gamma_error = estimated_amplitude_gamma - target_gamma
  )
```

```{r spectral-table}
knitr::kable(
  spectral_summary |>
    arrange(target_gamma, environment_replicate) |>
    mutate(
      across(
        c(
          estimated_power_exponent,
          estimated_amplitude_gamma,
          spectral_r_squared,
          expected_power_exponent,
          power_exponent_error,
          amplitude_gamma_error
        ),
        ~ round(.x, 4)
      )
    )
)
```

```{r spectral-plot, fig.width=8, fig.height=4}
spectral_summary |>
  ggplot(aes(x = target_gamma, y = estimated_amplitude_gamma)) +
  geom_abline(intercept = 0, slope = 1, linetype = 2) +
  geom_point(aes(colour = factor(environment_replicate)), size = 2) +
  labs(
    x = "Specified one_over_f_gamma",
    y = "Estimated amplitude exponent",
    colour = "Environment replicate"
  ) +
  theme_minimal()
```

## Example Spectra

The plot below shows fitted spectra for representative environment series.

```{r spectra-plot, fig.width=10, fig.height=8}
spectrum_data <- temperatures |>
  group_by(target_gamma) |>
  filter(environment_replicate == min(environment_replicate)) |>
  summarise(
    data = list({
      y <- temperature - mean(temperature)
      n <- length(y)
      fft_values <- stats::fft(y)
      frequencies <- seq_len(floor(n / 2)) / n
      power <- Mod(fft_values[seq_along(frequencies) + 1L])^2 / n
      tibble(frequency = frequencies, power = power)
    }),
    .groups = "drop"
  ) |>
  tidyr::unnest(data) |>
  filter(power > 0)

ggplot(spectrum_data, aes(x = frequency, y = power)) +
  geom_point(alpha = 0.08, size = 0.25) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, colour = "red") +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ target_gamma, ncol = 2) +
  labs(
    x = "Frequency",
    y = "Power"
  ) +
  theme_minimal()
```
