---
title: "Single-Community Walkthrough: Consumer-Resource"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single-Community Walkthrough: Consumer-Resource}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Purpose

This walkthrough defines one consumer-resource community, exposes it to one
temperature time series, runs the continuous-time consumer-resource simulator,
and makes a few basic plots.

## Set up R

```{r setup}
library(dplyr)
library(ggplot2)
library(tidyr)
library(community.simulator)

theme_set(theme_minimal(base_size = 12))
```

## Define One Environment

```{r environment, fig.alt="Line plot of temperature through time for the example environment."}
set.seed(3)
duration <- 120
time <- seq_len(duration)
temperature <- 20 +
  2 * sin(2 * pi * time / 50) +
  stats::arima.sim(model = list(ar = 0.6), n = duration, sd = 0.5)

environment <- tibble(
  time = time,
  temperature = as.numeric(temperature)
)

ggplot(environment, aes(x = time, y = temperature)) +
  geom_line(linewidth = 0.6) +
  labs(x = "Time", y = "Temperature", title = "Temperature Environment")
```

## Define One Community

```{r community}
community <- build_CR_community(
  S = 4,
  uptake_maximum_mean = 0.08,
  uptake_maximum_range = 0.00,
  uptake_maximum_distribution = "regular",
  uptake_optimum_mean = 20,
  uptake_optimum_range = 7,
  uptake_optimum_distribution = "regular",
  uptake_width_mean = 6,
  uptake_width_range = 1.5,
  uptake_width_distribution = "regular",
  half_saturation_mean = 80,
  half_saturation_range = 10,
  half_saturation_distribution = "regular",
  consumer_death_rate = 0.03,
  resource_renewal_rate = 1,
  resource_supply = 800,
  conversion_efficiency = 1,
  resource_use_mode = "shared_to_private",
  active_resource = 1,
  private_resource_use_distribution = "beta",
  private_resource_use_mean = 0.7,
  private_resource_use_precision = 18,
  community_seed = 13
)
```

### Consumer traits

```{r traits}
consumer_traits <- tibble(
  species = paste0("Spp", seq_len(community$S)),
  max_uptake = community$uptake_maximum_i,
  uptake_optimum = community$uptake_optimum_i,
  uptake_width = community$uptake_width_i,
  death_rate = community$d_i,
  private_resource_fraction = community$private_resource_use_i,
  shared_resource_fraction = 1 - community$private_resource_use_i
)

consumer_traits
```

### Temperature uptake curves

```{r uptake-curves, fig.alt="Line plot of approximate high-resource consumer growth against temperature for each species."}
temperature_grid <- tibble(temperature = seq(5, 35, length.out = 200))

uptake_curves <- tidyr::crossing(
  consumer_traits,
  temperature_grid
) |>
  mutate(
    uptake_capacity = max_uptake *
      exp(-0.5 * ((temperature - uptake_optimum) / uptake_width)^2),
    expected_growth_at_high_resource = uptake_capacity - death_rate
  )

ggplot(uptake_curves, aes(x = temperature, y = expected_growth_at_high_resource, colour = species)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey50") +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Temperature",
    y = "Approximate growth at high resource",
    title = "Consumer Thermal Uptake Curves"
  )
```

### Resource-use matrix

```{r resource-use, fig.alt="Heatmap of consumer resource-use fractions."}
resource_use <- as.data.frame(as.table(community$resource_use_ij)) |>
  rename(consumer = Var1, resource = Var2, use_fraction = Freq)

ggplot(resource_use, aes(x = resource, y = consumer, fill = use_fraction)) +
  geom_tile(colour = "white") +
  scale_fill_viridis_c(limits = c(0, 1)) +
  coord_equal() +
  labs(
    x = "Resource",
    y = "Consumer",
    fill = "Use fraction",
    title = "Resource-Use Matrix"
  )
```

## Run Dynamics

```{r dynamics}
initial_consumer_abundances <- rep(12, community$S)
initial_resource_values <- rep(600, community$R)

simulation <- simulator_consumer_resource_continuous(
  input_com_params = community,
  TcelSeries = matrix(environment$temperature, nrow = 1),
  initial_consumer_abundances = initial_consumer_abundances,
  initial_resource_values = initial_resource_values,
  times = environment$time,
  output_times = environment$time,
  temperature_interpolation = "linear",
  consumer_immigration_rate = 0.01,
  ode_method = "lsoda",
  rtol = 1e-6,
  atol = 1e-8,
  max_step = 1
)

consumers <- simulation$consumers |>
  mutate(time = environment$time) |>
  pivot_longer(
    cols = starts_with("Spp"),
    names_to = "species",
    values_to = "abundance"
  )

resources <- simulation$resources |>
  mutate(time = environment$time) |>
  pivot_longer(
    cols = starts_with("Res"),
    names_to = "resource",
    values_to = "amount"
  )
```

## Plot Dynamics

```{r plot-dynamics, fig.alt="Line plot of consumer abundance through time for each species."}
total_abundance <- consumers |>
  group_by(time) |>
  summarise(total_abundance = sum(abundance), .groups = "drop")

ggplot(consumers, aes(x = time, y = abundance, colour = species)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time", y = "Consumer abundance", title = "Consumer Dynamics")
```

```{r resource-dynamics, fig.alt="Line plot of resource amounts through time for each resource."}
ggplot(resources, aes(x = time, y = amount, colour = resource)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time", y = "Resource amount", title = "Resource Dynamics")
```

```{r total-abundance, fig.alt="Line plot of total consumer abundance through time."}
ggplot(total_abundance, aes(x = time, y = total_abundance)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time", y = "Total consumer abundance", title = "Community Total Abundance")
```

## Interactive Shiny Version

You can explore the same kind of one-community setup interactively with the
Shiny simulation explorer:

```{r shiny-explorer, eval = FALSE}
run_simulation_explorer()
```

In the app, choose **Consumer-resource**, press **Build community** to inspect
the resource-use matrix, uptake curves, and binary viability-based community
performance curve, and press **Simulate** only when you want to run the ODE
dynamics. The app has both a simple specification mode and a detailed mode for
expanding the CR resource-use and resource-supply parameters.
