---
title: "Getting Started with a Small Experiment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with a Small Experiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Overview

This guide runs a very small end-to-end experiment using
`community.simulator`. The goal is to show the simplest successful workflow for
creating an experiment folder, running the standard workflow, and inspecting the
main summary output.

The example:

- creates a project folder and an experiment subfolder
- copies a bundled experiment-definition file into the experiment folder
- highlights the point where a user would normally edit the YAML design
- runs the standard workflow with `run_experiment()`
- inspects the saved summary output

```{r setup}
library(community.simulator)
```

## Run a Tiny Example

```{r run-example}
project_folder_location <- tempdir()
experiment_name <- "getting_started_example"
experiment_design_filename <- "experiment.yaml"
experiment_folder <- file.path(project_folder_location, experiment_name)

dir.create(experiment_folder, recursive = TRUE, showWarnings = FALSE)
file.copy(
  system.file(
    "experiment_templates/lv_discrete.yaml",
    package = "community.simulator"
  ),
  file.path(experiment_folder, experiment_design_filename),
  overwrite = TRUE
)

## In a real project, this is the point where you would edit the YAML file
## inside `experiment_folder` before running the workflow.

outputs <- suppressWarnings(
  suppressMessages(
    run_experiment(
      experiment_folder_location = project_folder_location,
      experiment_name = experiment_name,
      experiment_design_filename = experiment_design_filename,
      overwrite = TRUE,
      verbose = FALSE,
      confirm_run = FALSE
    )
  )
)

outputs
```

## Inspect the Main Output

```{r inspect-output}
community_measures <- readRDS(outputs$community_measures)

community_measures[, c("case_id", "richness", "temperature_mean", "cv_total_abundance")]
```

## Inspect the Experiment Log

Each run writes a plain-text log file. The log is newline-delimited JSON: each
line is a complete record that can be read by humans or parsed by software.

```{r inspect-log}
readLines(outputs$experiment_log, n = 5)
```

## Main Output Files

```{r output-files}
names(outputs)
```

The most important outputs are:

- `experiment_table.RDS`: one row per simulation case
- `experiment_log.txt`: experiment specification, workflow events, and timings
- `temperatures.db`: environmental time series
- `dynamics.db`: saved population dynamics, when enabled
- `community_measures.RDS`: case-level summary measures

## Next Steps

Once this small example works, the usual next steps are:

- create a project folder that contains multiple experiment subfolders
- copy an example YAML file into each experiment folder, then edit it to change
  species, temperature, or replicate settings
- rerun the workflow for a new experiment name
- inspect `community_measures.RDS`, `experiment_log.txt`, `dynamics.db`, and
  `temperatures.db`
- use the single-community walkthroughs in `reports/examples` to understand how
  each model behaves before scaling up to experiment grids
