---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette runs a very small end-to-end experiment using
`community.simulator`. The goal is to show the simplest successful workflow for
new users.

The example:

- creates a project folder and an experiment subfolder
- copies a bundled experiment-definition file into the experiment folder
- highlights the point where a user would normally edit the JSON design
- runs the standard workflow with `run_experiment()`
- inspects the saved summary output

```{r setup}
library(community.simulator)
```

## Run a Tiny Example

```{r}
project_folder_location <- tempdir()
experiment_name <- "getting_started_example"
experiment_design_filename <- "experiment_definition_template_v0.7.json"

setup <- setup_example_experiment(
  experiment_folder_location = project_folder_location,
  experiment_name = experiment_name,
  example_experiment_name = "test_experiment1",
  experiment_design_filename = experiment_design_filename,
  verbose = FALSE
)

## In a real project, this is the point where you would edit the JSON file
## inside `setup$experiment_folder` before running the workflow.

outputs <- suppressWarnings(
  suppressMessages(
    run_experiment(
      experiment_folder_location = project_folder_location,
      experiment_name = experiment_name,
      experiment_design_filename = experiment_design_filename,
      overwrite = TRUE,
      verbose = FALSE
    )
  )
)

outputs
```

## Inspect the Main Output

```{r}
community_measures <- readRDS(outputs$community_measures)

community_measures[, c("case_id", "richness", "temperature_mean", "CV_totab")]
```

## Next Steps

Once this small example works, the usual next steps are:

- create a project folder that contains multiple experiment subfolders
- copy an example JSON file into each experiment folder, then edit it to change
  species, temperature, or replicate settings
- rerun the workflow for a new experiment name
- inspect `community_measures.RDS`, `dynamics.db`, and `temperatures.db`
- use `make_plots_for_one_community()` for case-level plots
