---
title: "Single-Community Walkthrough: Continuous-Time LV"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single-Community Walkthrough: Continuous-Time LV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Purpose

This walkthrough defines one Lotka-Volterra community, exposes it to one
temperature time series, runs the continuous-time simulator, and makes a few
basic plots.

## Interactive Shiny Version

You can explore the same kind of one-community setup interactively with the
Shiny simulation explorer:

```{r shiny-explorer, eval = FALSE}
run_simulation_explorer()
```

In the app, choose **Continuous-time LV**, press **Build community** to inspect
the interaction matrix and thermal performance curves, and press **Simulate**
only when you want to run the ODE dynamics. The app has both a simple
specification mode and a detailed mode for expanding the LV interaction
treatment and continuous-time controls.

```{r setup}
library(dplyr)
library(ggplot2)
library(tidyr)
library(community.simulator)

theme_set(theme_minimal(base_size = 12))
```

## Define One Environment

```{r environment, fig.alt="Line plot of temperature through time for the example environment."}
set.seed(2)
duration <- 120
time <- seq_len(duration)
temperature <- 20 +
  2.5 * sin(2 * pi * time / 45) +
  stats::arima.sim(model = list(ar = 0.55), n = duration, sd = 0.6)

environment <- tibble(
  time = time,
  temperature = as.numeric(temperature)
)

ggplot(environment, aes(x = time, y = temperature)) +
  geom_line(linewidth = 0.6) +
  labs(x = "Time", y = "Temperature", title = "Temperature Environment")
```

## Define One Community

```{r community}
community <- make_a_community(
  S = 4,
  a_b_mean = 0.32,
  a_b_range = 0.00,
  a_b_distribution = "regular",
  b_opt_mean = 20,
  b_opt_range = 7,
  b_opt_distribution = "regular",
  sd_perf_distribution = "regular",
  sd_perf_mean = 7.5,
  sd_perf_range = 2,
  community_seed = 12,
  a_d = 0.02,
  z = 0.03,
  lv_interaction_spec = list(
    label = "symmetric_competition",
    type = "competition",
    symmetry = "symmetric",
    distribution = "uniform",
    parameters = list(min = 0.03, max = 0.16),
    diagonal = 1
  )
)
```

### Species traits

```{r traits}
species_traits <- tibble(
  species = paste0("Spp", seq_len(community$S)),
  max_birth = community$a_b_i,
  thermal_optimum = community$b_opt_i,
  width = community$sd_perf_i,
  death_intercept = community$a_d_i,
  death_temperature_slope = community$z_i
)

species_traits
```

### Temperature performance curves

```{r performance-curves, fig.alt="Line plot of intrinsic growth rate against temperature for each species."}
temperature_grid <- tibble(temperature = seq(5, 35, length.out = 200))

performance_curves <- tidyr::crossing(
  species_traits,
  temperature_grid
) |>
  mutate(
    birth = max_birth * exp(-0.5 * ((temperature - thermal_optimum) / width)^2),
    death = death_intercept * exp(death_temperature_slope * temperature),
    intrinsic_growth = birth - death
  )

ggplot(performance_curves, aes(x = temperature, y = intrinsic_growth, colour = species)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey50") +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Temperature",
    y = "Intrinsic growth rate",
    title = "Species Thermal Performance Curves"
  )
```

### Interaction matrix

```{r interaction-matrix, fig.alt="Heatmap of the Lotka-Volterra interaction matrix."}
interaction_matrix <- as.data.frame(as.table(community$alpha_ij)) |>
  rename(consumer = Var1, neighbour = Var2, interaction = Freq)

ggplot(interaction_matrix, aes(x = neighbour, y = consumer, fill = interaction)) +
  geom_tile(colour = "white") +
  scale_fill_viridis_c() +
  coord_equal() +
  labs(
    x = "Neighbour species",
    y = "Focal species",
    fill = "alpha",
    title = "LV Interaction Matrix"
  )
```

## Run Dynamics

```{r dynamics}
initial_abundances <- rep(25, community$S)

abundances <- simulator_lv_continuous(
  input_com_params = community,
  TcelSeries = matrix(environment$temperature, nrow = 1),
  initial_abundances = initial_abundances,
  times = environment$time,
  output_times = environment$time,
  temperature_interpolation = "linear",
  immigration_rate = 0.05,
  immigration_mode = "continuous",
  ode_method = "lsoda",
  rtol = 1e-6,
  atol = 1e-8,
  max_step = 1
) |>
  mutate(time = environment$time) |>
  pivot_longer(
    cols = starts_with("Spp"),
    names_to = "species",
    values_to = "abundance"
  )
```

## Plot Dynamics

```{r plot-dynamics, fig.alt="Line plot of abundance through time for each species."}
total_abundance <- abundances |>
  group_by(time) |>
  summarise(total_abundance = sum(abundance), .groups = "drop")

ggplot(abundances, aes(x = time, y = abundance, colour = species)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time", y = "Abundance", title = "Population Dynamics")
```

```{r total-abundance, fig.alt="Line plot of total community abundance through time."}
ggplot(total_abundance, aes(x = time, y = total_abundance)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time", y = "Total abundance", title = "Community Total Abundance")
```
