---
title: "User Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(community.simulator)
```

## Purpose

`community.simulator` is an R package for designing and running simulation
experiments on multispecies communities with temperature-dependent vital rates.
It was developed to explore how species traits, environmental variability, and
response diversity influence community dynamics and stability.

This guide is the package-level reference for users and reviewers. It explains
the model families, the experiment workflow, the YAML experiment specification,
and the main outputs. For executable, one-community examples, start with the
single-community walkthroughs. For a small end-to-end experiment, see the
getting-started vignette. For bundled experiment specifications, see the YAML
templates article, especially the richer `*_rich.yaml` examples when designing
multi-treatment experiments.

## Main Ideas

The package separates a simulation study into four layers:

- **Community specification**: species traits, interaction structure, or
  consumer-resource parameters.
- **Environment specification**: environmental temperature time series.
- **Experiment design**: the grid of community treatments, environment
  treatments, and replicates.
- **Workflow execution**: creating the experiment table, generating
  environments, simulating dynamics, and calculating summary measures.

The central workflow is controlled by a YAML experiment-definition file. The
package reads that file, expands treatment values into an experiment table, and
runs one simulation case for each row of that table.

## Current API

Most users only need the YAML workflow:

| Task | Function |
| --- | --- |
| Read and validate a YAML specification | `read_experiment_spec()` |
| Expand a YAML specification into cases | `create_experiment_table_from_spec()` |
| Run the full experiment workflow | `run_experiment()` |
| Run workflow stages manually | `create_environments_from_spec()`, `simulate_dynamics_from_spec()`, `get_community_measures_from_spec()` |
| Build one community from explicit parameters | `build_LV_community()`, `build_CR_community()` |
| Build one community from a resolved specification | `build_community_from_spec()`, `build_LV_community_from_spec()`, `build_CR_community_from_spec()` |
| Explore one simulation interactively | `run_simulation_explorer()` |
| Make standard diagnostic plots | `plot_case_temperature()`, `plot_case_abundances()`, `plot_case_total_abundance()`, `plot_community_matrix()`, `plot_resource_dynamics()` |

The direct simulator functions remain available for advanced use, but the
recommended experiment interface is the YAML specification plus
`run_experiment()`.

For interactive exploration of one community, launch the Shiny app with
`run_simulation_explorer()`. The app is documented in the Shiny simulation
explorer article on the pkgdown site.

## Supported Model Families

The package currently supports three model families.

| Model family | `model.type` | Main use |
| --- | --- | --- |
| Discrete-time Lotka-Volterra | `"lv_discrete"` | Fast experiments with temperature-dependent intrinsic growth and LV interaction matrices |
| Continuous-time Lotka-Volterra | `"lv_continuous"` | ODE-based LV dynamics with temperature interpolation and immigration controls |
| Continuous-time consumer-resource | `"consumer_resource_continuous"` | Resource-mediated consumer dynamics with temperature-dependent uptake |

Set `model.type` in the YAML file to choose the model family. If `model.type`
is omitted, the package uses the discrete-time Lotka-Volterra model.

## Population Dynamic Models

The package has three population dynamic models. They share the same broad
workflow, but differ in how species interact and how abundance changes through
time.

| Model | State variables | Main ecological structure | Time update |
| --- | --- | --- | --- |
| Discrete-time LV | Species abundances | Direct species interactions through an LV interaction matrix | Difference-equation update with temperature-dependent intrinsic growth |
| Continuous-time LV | Species abundances | Direct species interactions through an LV interaction matrix | ODE integration with temperature interpolation and optional immigration |
| Continuous-time consumer-resource | Consumer abundances and resource concentrations | Indirect interactions through shared and private resources | ODE integration of consumer and resource dynamics |

### Discrete-Time LV

The discrete-time LV model updates species abundances one time step at a time.
Temperature affects each species through its intrinsic growth rate, and the LV
interaction matrix determines how species affect one another within each update.
This model is useful for fast exploratory experiments and for cases where a
discrete-time approximation is adequate.

Select it with:

```yaml
model:
  type: lv_discrete
```

### Continuous-Time LV

The continuous-time LV model uses the same temperature-dependent species traits
and LV interaction matrix, but integrates the dynamics with `deSolve::ode()`.
This makes it useful when the continuous-time formulation is more natural, when
solver tolerances matter, or when continuous immigration is part of the model.

Select it with:

```yaml
model:
  type: lv_continuous
```

### Continuous-Time Consumer-Resource

The consumer-resource model simulates consumer abundances and resource
concentrations together. Consumers have temperature-dependent uptake curves, and
their interactions arise indirectly through resource use. Depending on
`resource_use_mode`, consumers can use one common resource, species-specific
resources, or a mix of shared and private resources.

Select it with:

```yaml
model:
  type: consumer_resource_continuous
```

## How Species Traits Are Specified

Species traits in a community are not usually specified by giving a separate
value for each trait of each species. Instead, an experiment usually defines
the distribution from which each species-level trait is generated.

This is why many trait fields appear as a set of three related specifications:

- a **mean**
- a **range**
- a **distribution**

The **mean** is the central value of the trait distribution. For example, if
`birth_optimum_mean` is `20`, the birth-rate thermal optima are centred around
20 degrees.

The **range** is the total spread of trait values around the mean. A range of
`4` means that generated values are drawn or placed across
`mean +/- 0.5 * range`. For a mean of `20` and a range of `4`, the trait values
come from the interval 18 to 22.

The **distribution** defines how species-level values are chosen within that
range. The two main trait distributions are:

- `"regular"`: species are assigned evenly spaced trait values across the
  interval.
- `"random_uniform"`: species trait values are drawn independently from a
  uniform distribution across the interval.

This distribution-based specification is useful for experiments because it lets
the user vary community-level properties, such as average thermal optimum or
among-species trait spread, without manually writing every species trait value.
Each community replicate receives a seed, so randomly generated trait values are
reproducible.

## Temperature-Dependent Performance

Both LV model variants use a Gaussian birth-rate curve and an exponential
death-rate curve. For species \(i\) at temperature \(T\):

\[
b_i(T) = a_{b,i} \exp\left[-\frac{1}{2}
  \left(\frac{T - b_{\mathrm{opt},i}}{sd_{\mathrm{perf},i}}\right)^2
\right]
\]

\[
d_i(T) = a_{d,i} \exp(z_i T)
\]

The intrinsic growth rate is:

\[
r_i(T) = b_i(T) - d_i(T)
\]

The width parameter is the standard-deviation scale of the Gaussian performance
curve. In YAML experiment files, the distribution of LV birth-curve widths is
controlled by:

- `birth_width_distribution`
- `birth_width_mean`
- `birth_width_range`

The consumer-resource model uses the same Gaussian width convention for uptake
curves. The corresponding YAML fields are:

- `uptake_width_distribution`
- `uptake_width_mean`
- `uptake_width_range`

## Lotka-Volterra Models

### Community Traits

LV communities are created from YAML experiment specifications by
`build_LV_community_from_spec()`. At the experiment level, the main trait
treatments are:

- `richness`: number of species in each community.
- `birth_maximum_mean`, `birth_maximum_range`,
  `birth_maximum_distribution`: maximum birth-rate heights.
- `birth_optimum_mean`, `birth_optimum_range`, `birth_optimum_distribution`:
  thermal optima.
- `birth_width_distribution`, `birth_width_mean`, `birth_width_range`:
  Gaussian performance-curve widths.
- `death_intercept`: death-rate intercept, shared across species.
- `death_temperature_slope`: exponential temperature sensitivity of death rate,
  shared across species.

Trait distributions currently support:

- `"regular"`: evenly spaced values across `mean +/- 0.5 * range`.
- `"random_uniform"`: independent uniform draws across the same interval.

Each unique community treatment gets a reproducible `community_seed` generated
from the experiment-level `random_seed`.

### Interaction Matrices

LV interactions are specified with `interactions.treatments`. This is a list of
named interaction treatments. Each treatment can include:

- `label`: readable treatment name used in the experiment table.
- `type`: `"none"`, `"competition"`, `"any"`, or `"predator_prey"`.
- `symmetry`: `"asymmetric"`, `"symmetric"`, or `"antisymmetric"`.
- `distribution`: `"constant"`, `"uniform"`, `"normal"`, `"lognormal"`, or
  `"gamma"`.
- `parameters`: distribution-specific numeric parameters.
- `diagonal`: diagonal value, usually `1`.

For example:

```yaml
interactions:
  treatments:
    - label: weak_asymmetric_competition
      type: competition
      symmetry: asymmetric
      distribution: uniform
      parameters:
        min: 0
        max: 0.2
      diagonal: 1
```

For `type = "competition"`, off-diagonal effects are positive. This is the
recommended way to ensure that all pairwise LV interactions are competitive.

### Discrete-Time LV Dynamics

The discrete-time LV simulator is `simulator_lv()`. It updates species
abundances through time using temperature-dependent intrinsic growth and the LV
interaction matrix.

This model is the default when `model.type` is absent or set to
`"lv_discrete"`.

### Continuous-Time LV Dynamics

The continuous-time LV simulator is `simulator_lv_continuous()`. It uses
`deSolve::ode()` and adds ODE-specific controls:

- `temperature_interpolation`: `"linear"` or `"constant"`.
- `immigration_rate`: immigration rate per species.
- `immigration_mode`: `"continuous"` or `"pulse"`.
- `ode_method`: ODE solver method passed to `deSolve::ode()`.
- `ode_rtol`, `ode_atol`: relative and absolute tolerances.
- `ode_max_step`: maximum solver step.
- `blowup_threshold`: abundance threshold above which the run stops.

Use `model.type: lv_continuous` in the YAML file to select this model.

## Consumer-Resource Model

The consumer-resource model is created by
`build_CR_community_from_spec()` and simulated by
`simulator_consumer_resource_continuous()`.

Consumers have temperature-dependent uptake curves:

\[
u_i(T) = u_{\max,i} \exp\left[-\frac{1}{2}
  \left(\frac{T - u_{\mathrm{opt},i}}{sd_{u,i}}\right)^2
\right]
\]

The main consumer-resource trait treatments are:

- `uptake_maximum_mean`, `uptake_maximum_range`,
  `uptake_maximum_distribution`:
  maximum uptake rates.
- `uptake_optimum_mean`, `uptake_optimum_range`,
  `uptake_optimum_distribution`:
  thermal optima for uptake.
- `uptake_width_distribution`, `uptake_width_mean`, `uptake_width_range`:
  uptake-curve widths.
- `half_saturation_mean`, `half_saturation_range`,
  `half_saturation_distribution`: Monod half-saturation constants.
- `consumer_death_rate`: consumer death rate.
- `resource_renewal_rate`: resource renewal rate.
- `resource_supply`: resource supply concentration.
- `conversion_efficiency`: conversion from uptake to consumer growth.

Use `model.type: consumer_resource_continuous` in the YAML file to select this
model.

### Resource-Use Modes

The consumer-resource model supports three resource-use modes:

- `"one_resource_all_consumers"`: all consumers use one active resource.
- `"diagonal"`: each consumer uses its corresponding resource.
- `"shared_to_private"`: each consumer splits uptake between one shared
  resource and a species-specific private resource.

For `"shared_to_private"`, the number of resources is `S + 1`. The
`active_resource` is the shared resource. Each species gets a private-resource
use fraction; its shared-resource fraction is one minus that private-resource
fraction. This is stored internally as `resource_specialization_i`.

Species-level private-resource use is controlled by:

- `private_resource_use_distribution`
- `private_resource_use_mean`
- `private_resource_use_range`
- `private_resource_use_precision`

Supported distributions are:

- `"constant"`: all species have the same specialization value.
- `"regular"`: evenly spaced values across `mean +/- 0.5 * range`.
- `"random_uniform"`: independent uniform draws across that interval.
- `"beta"`: beta-distributed species values with mean
  `private_resource_use_mean` and precision
  `private_resource_use_precision`.

Higher beta precision gives less among-species variation around the mean.

## Environment Specification

Environmental variation is generated as temperature time series. The current
experiment workflow uses a `1/f` process from the `primer` package. The main
temperature fields are:

- `temperature_mean`: mean temperature.
- `temperature_sd`: standard deviation of the temperature series.
- `one_over_f_gamma`: slope of the frequency-power relationship.
- `number_of_environment_replicates`: number of environmental replicates.
- `environment_sharing`: how temperature series are shared among cases.

The main `environment_sharing` options are:

- `"same_per_replicate"`: cases with the same temperature treatment and
  replicate number share an environmental time series.
- `"all_different"`: each simulation case gets its own environmental time
  series.

The duration of each run is controlled by:

- `burn_in_duration`
- `experiment_duration`

The workflow creates temperature series long enough to include both burn-in and
experiment periods.

## Experiment Workflow

The usual workflow is:

1. Create a project folder that can contain multiple experiment folders.
2. Create an experiment folder.
3. Create or copy an experiment-definition YAML file into the experiment folder.
4. Edit the YAML file to define model, community, environment, and replicate
   treatments.
5. Create an R script in the experiment folder. Use it to run the experiment with `run_experiment()`.
6. Inspect outputs in the experiment folder.

A compact example is:

```{r workflow-example, eval = FALSE}
project_folder_location <- file.path("~/Desktop", "community_simulator_projects")
experiment_name <- "discrete_lv_example"
experiment_design_filename <- "experiment.yaml"
experiment_folder <- file.path(project_folder_location, experiment_name)

dir.create(experiment_folder, recursive = TRUE, showWarnings = FALSE)
file.copy(
  system.file(
    "experiment_templates/lv_discrete.yaml",
    package = "community.simulator"
  ),
  file.path(experiment_folder, experiment_design_filename),
  overwrite = TRUE
)

outputs <- run_experiment(
  experiment_folder_location = project_folder_location,
  experiment_name = experiment_name,
  experiment_design_filename = experiment_design_filename,
  overwrite = FALSE,
  confirm_run = TRUE
)
```

The high-level wrapper runs the main steps in order:

- `read_experiment_spec()`
- `create_experiment_table_from_spec()`
- `create_environments_from_spec()`
- `simulate_dynamics_from_spec()`
- `get_community_measures_from_spec()`

These functions can also be run individually when debugging or developing a new
workflow. By default, workflow functions stop if an output file already exists.
Use `overwrite = TRUE` only when you deliberately want to replace existing
outputs.

## Interactive Shiny Explorer

The package includes a small Shiny app for interactively exploring one
community at a time. It is intended as a learning and reviewer aid rather than
a replacement for the YAML experiment workflow.

Launch the app from an installed package with:

```{r shiny-run, eval = FALSE}
run_simulation_explorer()
```

If you are working from the RStudio project in a local checkout, you can also
source the project launcher:

```{r shiny-run-local, eval = FALSE}
source("scripts/run_simulation_explorer.R")
```

The app supports all three population dynamic models:

- discrete-time LV
- continuous-time LV
- continuous-time consumer-resource

The app separates community construction from simulation. Press **Build
community** to create one community and inspect its structure, traits, thermal
performance curves, and community performance curve. Press **Simulate** only
when you want to run the dynamics. If a simulation is running for too long,
press **Stop simulation** to interrupt the background process.

The app has two specification modes:

- **Simple**: a compact set of controls for quick exploration.
- **Detailed**: model-specific controls, including LV interaction type,
  symmetry, distribution, and diagonal value, and CR resource-use mode,
  private-resource use distribution, half-saturation, resource supply, and
  related parameters.

For the consumer-resource model, the community performance curve is shown as a
binary viability summary: at each temperature it gives the fraction of
consumers with positive expected growth.

## Bundled Example Designs

The package includes compact and richer example YAML templates under
`inst/experiment_templates`.

The compact templates are intentionally small, so they are useful as smoke
tests and as starting points for the smallest possible experiment:

- `lv_discrete.yaml`
- `lv_continuous.yaml`
- `consumer_resource.yaml`

The richer templates show a more reviewer- or user-facing experiment
specification, including treatments, replicate structure, interaction or
resource-use options, output controls, and parallel settings:

- `lv_discrete_rich.yaml`
- `lv_continuous_rich.yaml`
- `consumer_resource_rich.yaml`

Copy one into an experiment folder, rename it if useful, and edit it before
running `run_experiment()`. The richer templates are still modest examples, but
they are closer to the shape of an experiment a user would adapt for analysis.

## YAML File Conventions

Experiment YAML files are read by `read_experiment_spec()`. They are declarative:
values are written directly as YAML scalars, lists, and mappings. The package
does not evaluate R expressions from the experiment specification.

Single numeric and character values are written directly:

```yaml
model:
  type: lv_continuous

environment:
  temperature:
    mean: 20
```

Treatment vectors are written as YAML lists. If `treatments.mode` is omitted,
the package uses full factorial expansion.

```yaml
treatments:
  values:
    traits.birth_width.mean: [5, 10, 15]
    environment.temperature.sd: [0.5, 1, 2]
```

Paired treatments are written as explicit rows:

```yaml
treatments:
  mode: paired
  values:
    - traits.birth_width.mean: 5
      environment.temperature.sd: 0.5
    - traits.birth_width.mean: 10
      environment.temperature.sd: 1
```

Treatment paths must already exist in the baseline specification. This catches
spelling errors before the experiment starts.

### Common YAML Errors

`read_experiment_spec()` validates an experiment before the workflow starts. The
goal is to fail early, before environments or simulations are created.

Common validation errors include:

| Error type | Example message | How to fix it |
| --- | --- | --- |
| Missing section | `Experiment specification is missing required section(s): traits` | Add the missing top-level section to the YAML file. |
| Invalid model name | `` `model.type` must be one of: lv_discrete, lv_continuous, consumer_resource_continuous. `` | Use one of the supported model names exactly. |
| Wrong value type | `` `community.richness` must be an integer >= 1. `` | Use a single positive integer for counts and replicate numbers. |
| Invalid distribution | `` `traits.birth_width.distribution` must be one of: regular, random_uniform. `` | Use one of the distributions supported for that trait. |
| Invalid numeric range | `` `traits.birth_width.mean - 0.5 * traits.birth_width.range` must be positive. `` | Make the mean and range imply possible species-level values within the allowed range. |
| Invalid interaction treatment | `predator-prey interactions cannot use symmetric symmetry` | Use `asymmetric` or `antisymmetric` symmetry for predator-prey interactions. |
| Invalid private resource use | `` `resources.private_use.mean` must be greater than 0 and less than 1. `` | For beta-distributed private resource use, keep the mean strictly between 0 and 1. |
| Misspelled treatment path | `` Treatment path `traits.birth_wdith.mean` does not exist... Did you mean `traits.birth_width.mean`? `` | Correct the dotted path so it matches a field already present in the baseline YAML specification. |

## YAML Parameter Reference

This table describes the main public-facing YAML sections and fields used by
the bundled example experiments.

| Parameter | Models | Meaning and options |
| --- | --- | --- |
| `experiment.name` | All | Human-readable experiment name. |
| `experiment.random_seed` | All | Experiment-level random seed used to make generated communities and environments reproducible. |
| `model.type` | All | Population dynamic model. Options are `lv_discrete`, `lv_continuous`, and `consumer_resource_continuous`. |
| `community.richness` | All | Number of species or consumers in each community. |
| `community.replicates` | All | Number of replicate communities for each community-treatment combination. |
| `environment.replicates` | All | Number of replicate temperature series for each environmental treatment. |
| `environment.sharing` | All | How generated environments are shared among simulation cases. Options are `same_per_replicate` and `all_different`. |
| `environment.temperature.mean` | All | Mean of the generated environmental temperature time series. |
| `environment.temperature.sd` | All | Standard deviation of the generated environmental temperature time series. |
| `environment.temperature.one_over_f_gamma` | All | Slope parameter for the `1/f` environmental noise process. |
| `simulation.burn_in_duration` | All | Number of initial time steps or time units treated as burn-in. |
| `simulation.experiment_duration` | All | Number of post-burn-in time steps or time units used for the experiment. |
| `simulation.temperature_interpolation` | Continuous LV, CR | How the continuous-time simulator interpolates temperature between generated values. Options are `linear` and `constant`. |
| `simulation.immigration_rate` | Continuous LV | Immigration rate per species. |
| `simulation.immigration_mode` | Continuous LV | How immigration is applied. Options are `continuous` and `pulse`. |
| `simulation.consumer_immigration_rate` | CR | Immigration rate for consumers. |
| `simulation.initial_consumer_total_abundance` | CR | Initial total consumer abundance distributed across consumers. |
| `simulation.resource_initial_value` | CR | Initial concentration for each resource. |
| `simulation.ode.method` | Continuous LV, CR | ODE solver method passed to `deSolve::ode()`, commonly `lsoda`. |
| `simulation.ode.rtol` | Continuous LV, CR | Relative tolerance for the ODE solver. |
| `simulation.ode.atol` | Continuous LV, CR | Absolute tolerance for the ODE solver. |
| `simulation.ode.max_step` | Continuous LV, CR | Maximum ODE solver step size. |
| `simulation.blowup_threshold` | Continuous LV, CR | Abundance threshold above which a simulation is treated as having blown up. |
| `simulation.negative_tolerance` | CR | Small numerical tolerance used when checking negative values in ODE output. |
| `traits.birth_maximum.*` | LV | Mean, range, and distribution for maximum birth-rate performance. |
| `traits.birth_optimum.*` | LV | Mean, range, and distribution for birth-rate thermal optima. |
| `traits.birth_width.*` | LV | Mean, range, and distribution for Gaussian birth-rate performance-curve widths. |
| `traits.death.intercept` | LV | Intercept of the death-rate curve, shared across species. |
| `traits.death.temperature_slope` | LV | Temperature sensitivity of the death-rate curve, shared across species. |
| `interactions.treatments` | LV | List of LV interaction-matrix treatments. Each treatment can include `label`, `type`, `symmetry`, `distribution`, `parameters`, and `diagonal`. |
| `traits.uptake_maximum.*` | CR | Mean, range, and distribution for maximum consumer uptake. |
| `traits.uptake_optimum.*` | CR | Mean, range, and distribution for uptake thermal optima. |
| `traits.uptake_width.*` | CR | Mean, range, and distribution for Gaussian uptake-curve widths. |
| `traits.half_saturation.*` | CR | Mean, range, and distribution for Monod half-saturation constants. |
| `resources.use_mode` | CR | Resource-use structure. Options are `one_resource_all_consumers`, `diagonal`, and `shared_to_private`. |
| `resources.private_use.*` | CR | Distribution, mean, range or precision for species-level private-resource use. |
| `resources.consumer_death_rate` | CR | Consumer death rate, shared across consumers. |
| `resources.renewal_rate` | CR | Chemostat resource renewal rate. |
| `resources.supply` | CR | Resource supply concentration. |
| `resources.conversion_efficiency` | CR | Conversion efficiency from resource uptake to consumer growth. |
| `parallel.workers` | All | Number of worker processes for parallel steps. Use a positive integer, `available_cores`, `available_cores_minus_1`, or `auto` (`auto` is equivalent to `available_cores_minus_1`). |
| `parallel.environments` | All | Logical. If `TRUE`, generate environmental time series in parallel where supported. |
| `parallel.simulations` | All | Logical. If `TRUE`, simulate cases in parallel where supported. |
| `parallel.community_measures` | All | Logical. If `TRUE`, calculate community performance curve measures in parallel where supported. |
| `output.dynamics_save_every` | All | Interval between saved population-dynamics output time points. |
| `output.resources_save_every` | CR | Interval between saved resource output time points. |
| `output.save_dynamics` | All | Logical. If `TRUE`, write `dynamics.db`. |
| `output.save_resources` | CR | Logical. If `TRUE`, write `resources.db`. |
| `treatments.mode` | All | Treatment expansion mode. Defaults to `factorial`; use `paired` for explicit treatment rows. |
| `treatments.values` | All | Named dotted paths and treatment values to apply to the baseline specification. |

## Outputs

The workflow writes outputs inside the experiment folder. The core outputs are:

- `experiment_table.RDS`: one row per simulation case, including the community
  object for each case.
- `experiment_log.txt`: newline-delimited JSON records describing the
  experiment specification, workflow events, output paths, and elapsed time for
  each step.
- `temperatures.db`: SQLite database containing generated environmental
  temperature series.
- `dynamics.db`: SQLite database containing saved population dynamics, when
  dynamic output is enabled.
- `community_measures.RDS`: case-level summary measures.

Recent workflows also save compact summary files during simulation:

- `simulation_summaries.RDS`
- `population_summaries.RDS`

These compact summaries allow downstream community-measure calculations without
always reading full dynamics from SQLite.

## Community Measures

`get_community_measures_from_spec()` calculates case-level summary measures as
part of the YAML workflow. The exact columns depend on the model type and
available intermediate outputs, but common outputs include:

- community stability, including `cv_total_abundance`, the coefficient of
  variation of total abundance through time
- mean and standard deviation of total abundance
- environmental summaries, such as mean temperature
- realized performance-optimum summaries, such as
  `realized_mean_performance_optimum`; these are calculated from the sampled
  species traits in the community, so they can differ from the design-specified
  trait-distribution values
- community performance or viability summaries used in response-diversity
  analyses, such as `cv_community_performance_info` for LV models and
  `cv_community_viability_binary_info` for consumer-resource models

For most analyses, `community_measures.RDS` is the first file to inspect after
a successful experiment.

### Community-Measures Output Reference

The table below describes the main columns produced in `community_measures.RDS`.
Some columns are model-specific or depend on output settings, so not every
experiment will contain every measure.

| Column | Meaning |
| --- | --- |
| `case_id` | Unique simulation-case identifier. |
| `model_type` | Population dynamic model used for the case. |
| `treatment_id` | Identifier for the expanded YAML treatment combination. |
| `treatment_label` | Readable label for the expanded YAML treatment combination. |
| `interaction_treatment_label` | LV interaction-matrix treatment label; `NA` for CR models. |
| `community_replicate` | Community replicate number. |
| `environment_replicate` | Environment replicate number. |
| `env_series_id` | Identifier for the generated environmental temperature series. |
| `temperature_mean` | Design-specified mean temperature for the case. |
| `temperature_sd` | Design-specified standard deviation of environmental temperature. |
| `one_over_f_gamma` | Design-specified `1/f` environmental-noise slope. |
| `richness` | Number of species or consumers in the community. |
| `community_id` | Identifier for the realized community object. |
| `mean_total_abundance` | Mean total abundance through the post-burn-in experiment period. |
| `sd_total_abundance` | Standard deviation of total abundance through the post-burn-in experiment period. |
| `cv_total_abundance` | Coefficient of variation of total abundance; the main community-stability measure. |
| `comm_temperature_sensitivity` | Slope from a linear model of total abundance against temperature for the case. |
| `synchrony_abundance` | Loreau-style synchrony of population abundances, calculated from total-abundance variance relative to summed population standard deviations. |
| `mean_population_cv_abundance` | Mean population-level abundance CV, weighted by mean relative abundance. |
| `final_total_abundance` | Total abundance at the final saved time point. |
| `final_min_abundance` | Minimum species abundance at the final saved time point. |
| `final_mean_abundance` | Mean species abundance at the final saved time point. |
| `final_max_abundance` | Maximum species abundance at the final saved time point. |
| `final_richness_above_extinction_threshold` | Number of species above the extinction threshold at the final saved time point. |
| `n_extinct_final` | Number of species at or below the extinction threshold at the final saved time point. |
| `any_extinct_final` | Whether any species is extinct at the final saved time point. |
| `n_extinct_ever` | Number of species that crossed the extinction threshold at any saved time point. |
| `any_extinct_ever` | Whether any species crossed the extinction threshold at any saved time point. |
| `min_abundance_ever` | Minimum species abundance observed during the saved experiment period. |
| `extinction_threshold` | Threshold used for extinction summaries. |
| `performance_optimum_trait` | Trait used for performance-optimum summaries: `birth_optimum` for LV models and `uptake_optimum` for CR models. |
| `sum_relative_performance_optimum` | Sum across species of realized performance optimum minus mean environmental temperature. |
| `minimum_absolute_relative_performance_optimum` | Smallest absolute distance between a species' realized performance optimum and mean environmental temperature. |
| `realized_mean_performance_optimum` | Mean realized performance optimum across the sampled species in the community. The `realized` prefix indicates that this is calculated from the actual sampled community, not copied from the design-specified trait distribution. |
| `realized_sd_performance_optimum` | Standard deviation of realized performance optima across the sampled species in the community. |
| `cv_community_performance_info` | CV of the LV community performance curve evaluated over the actual experienced temperature series. |
| `cv_community_performance_naive` | CV of the LV community performance curve evaluated over an evenly spaced temperature sequence spanning the experienced range. |
| `cv_community_viability_binary_info` | CV of CR binary community viability evaluated over the actual experienced temperature series. |
| `cv_community_viability_binary_naive` | CV of CR binary community viability evaluated over an evenly spaced temperature sequence spanning the experienced range. |
| `cv_community_viability_soft_info` | CV of CR soft community viability evaluated over the actual experienced temperature series. |
| `cv_community_viability_soft_naive` | CV of CR soft community viability evaluated over an evenly spaced temperature sequence spanning the experienced range. |
| `synchrony_performance_info` | Synchrony of species performance curves evaluated over the actual experienced temperature series. |
| `synchrony_performance_naive` | Synchrony of species performance curves evaluated over an evenly spaced temperature sequence spanning the experienced range. |
| `mean_species_cv_performance_info` | Mean species-level performance CV, weighted by mean performance, over the actual experienced temperature series. |
| `mean_species_cv_performance_naive` | Mean species-level performance CV, weighted by mean performance, over an evenly spaced temperature sequence spanning the experienced range. |

## Parallel Processing and Runtime Reporting

Experiment YAML files can include runtime controls for larger runs:

- `parallel.environments`: generate environmental time series in parallel.
- `parallel.simulations`: simulate cases in parallel.
- `parallel.community_measures`: calculate community performance curve measures
  in parallel.
- `parallel.workers`: number of worker processes.
- `runtime.update_every`: how often progress updates are printed.
- `runtime.environment_progress`: whether environment-generation progress is
  printed.

When parallel generation or simulation is enabled, database writing is still
handled by the parent process. Parallel processing is intended for macOS/Linux.

The experiment log records elapsed time for the main workflow steps. It is
plain text newline-delimited JSON, so it is readable by humans and easy to parse
from R or other tools.
