---
title: "User_guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User_guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(community.simulator)

```

# Introduction

This `community.simulator` package is a tool to simulate the dynamics of a community of species and to design and run experiments where features of species and the environment can be manipulated.

It was originally created to allow investigation of how response diversity influences community stability.

## The population dynamic model

We have $S$ species that can be interacting and whose vital rates are temperature dependent. We assume density-dependent birth rate, ($B$), and death rate ($D$) in a discrete-time version of the classical Lotka--Volterra model [@de2013predicting; @vasseur2020impact] to get instantaneous growth rate, $\tilde{r}_{i}(t)$, for species, $i$, in year $t$:

```{=tex}
\begin{equation}\label{eq.r}
   \tilde{r}_{i}(t) = ln N_{i}(t+1) - ln N_{i}(t) = B(N_{i}(t),N_{j}(t),T(t)) - D(N_{i}(t),N_{j}(t),T(t))(\#eq:r)
\end{equation}
```
Here, $N(t)$ represents the biomass at year $t$, and $i,j$ are indices for two different species.

The per-capita birth and death rates for $i^{th}$ species are represented as:

```{=tex}
\begin{equation}\label{eq.B}
   B_{i} = b_{0,i}(T)-\beta (N_{i}+\sum_{i \neq j = 1}^{S} \alpha_{ij}N_{j})(\#eq:B)
\end{equation}
```
```{=tex}
\begin{equation}\label{eq.D}
   D_{i} = d_{0,i}(T)-\delta (N_{i}+\sum_{i \neq j = 1}^{S} \alpha_{ij}N_{j})(\#eq:D)
\end{equation}
```
where, $\beta$ and $\delta$ are density-dependent constants, $\alpha_{i,j}$ is the competition coefficient between species $i$ and $j$, and

```{=tex}
\begin{equation}\label{eq.b0}
   b_{0,i}(T) = a_{b} e^{-(T-b_{opt,i})^2/s_{i}} (\#eq:b0)
\end{equation}
```
```{=tex}
\begin{equation}\label{eq.d0}
   d_{0,i}(T) = a_{d} e^{z_{i}T} (\#eq:d0)
\end{equation}
```
with $a_{b}$, $a_{d}$ as intercepts, and for $i^{th}$ species, $b_{opt,i}$ is the temperature that optimizes birth rate, $s_{i}$ governs the breadth of the birth function, $z_{i}$ scales the effect of temperature (in °C) to mimic the Arrhenius relationship.

Substituting Eqs. (\ref{eq.B}) - (\ref{eq.d0}) in Eq. (\ref{eq.r}), we get the following,

```{=tex}
\begin{equation}\label{eq.r2}
   \tilde{r}_{i}(t) = r_{m,i} \left( 1-\dfrac{\sum_{i, j = 1}^{S} \alpha_{ij}N_{j}}{K_{i}} \right) (\#eq:r2)
\end{equation}
```
where, $\alpha_{ii} = 1$, $r_{m,i} = (b_{0,i} - d_{0,i})$ is the intrinsic (maximum) rate of increase and $K_{i} = r_{m,i}/(\beta+\delta)$ is the carrying capacity for the $i^{th}$ species, respectively.

At each time step there is a small amount of immigration into the system. This is implemented by adding 0.1 to all species' biomass at each time step. Done in the `simulator_lv` function.



## Model parameters

The following need to be specified for a community (in addition to the number of species):

```{r c4}
#| echo: false
#| eval: true
species_and_community_properties <- read.csv("species_and_community_properties.csv")
species_and_community_properties |> 
  kableExtra::kbl() |> 
  kableExtra::kable_styling()
```

# Designing and experiment

An experiment consists of a number of "cases". Each case consists of a combination of community composition and environmental variability. The community composition is defined by the species present and their properties. The environmental variability is a time series of environmental conditions that the community is exposed to. When a single community has multiple replicates, then each replicate has the same community composition but different environmental variability.

The features of the experiment are given in a json file.



## Creating environmental variability

Currently created by 1 over f process with *gamma* (slope of frequency-power relationship) given in the json file. The 1 over f process is created with the `one_over_f` function from the (`primer` package)[https://search.r-project.org/CRAN/refmans/primer/html/00Index.html]. 

Also given in the json file are:

* The mean of the environmental time series.
* The standard deviation of the environmental time series.
* How the environmental time series vary among cases.

Currently three options for how environmental time series vary among cases.

1) `same_per_replicate` replicates of the same number share the same environmental time series. E.g., `case1_rep1` and `case2_rep1` share the same time series.
2) `all_different` all environmental time series are different



# Typical experiment workflow

The typical workflow is as follows:

1 Design an experiment.
2 Make the folder where the experiment information will be stored.
3 Run the experiment.
4 Make some calculation on the results.
5 Make some graphs of the results.

## An example experiment

You can find an example of running this workflow in the `test_experiments` folder of the package. It could be useful to make a copy of this folder and its contents somewhere convienent, for example on your desktop, using the following code:

```{r}
#| eval = FALSE
folder_to_copy <- system.file("test_experiments", package = "community.simulator")
folder_to_copy_to <- file.path("~/Desktop", "test_experiments")
dir.create(folder_to_copy_to)
file.copy(from = folder_to_copy, to = folder_to_copy_to, recursive = TRUE)

```

In the `test_experiments` folder you will find a file `test_experiment1.r` which goes through the process of designing, running, and analysing an experiment.

# JSON file specification

The design of an experiment is given in a JSON file.

An example of an experiment design file is in the `test_experiments` folder of the package. You can copy this file to your desktop (or another location if you change the value of `file_to_copy_to`):

```{r}
#| eval = FALSE
path_to_example_experiment_design_file <- system.file(file.path("test_experiments",
                      "test_experiment1",
                      "experiment_definition_template_v0.3.json"),
            package = "community.simulator")
file_to_copy_to <- file.path("~/Desktop", "experiment_definition_template_v0.3.json")
file.copy(from = path_to_example_experiment_design_file, to = folder_to_copy_to)

```


### Allowed values

All values in that file are parsed and then must be R expressions that can be evaluated.

**Single numbers**

"name": 0

```{r}
eval(parse(text = 0))
```

**Multiple numbers and functions**

"name": "c(0, 1, 2)"

```{r}
eval(parse(text = "c(0, 1, 2)"))
```

"name": "0:10"

```{r}
eval(parse(text = "0:10"))
```

"name": "seq(0, 10, 2)"

```{r}
eval(parse(text = "seq(0, 10, 2)"))
```



**Single words**

Do not use: "name": "word"

```{r eval = FALSE}
eval(parse(text = "word"))
```

It must be:

"name": "\"word\""

```{r}
eval(parse(text = "\"word\""))
```

**Multiple words**

"name": "c(\"word1\", \"word2\")"

```{r}
eval(parse(text = "c(\"word1\", \"word2\")"))
```


