---
title: "User_guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User_guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(community.simulator)

```

# Introduction

> Draft note: this user guide is still a work in progress. Some parts may change as the package interface and documentation continue to improve.

This `community.simulator` package is a tool to simulate the dynamics of a community of species and to design and run experiments where features of species and the environment can be manipulated.

It was originally created to allow investigation of how response diversity influences community stability.

## The population dynamic model

We have $S$ species that can be interacting and whose vital rates are temperature dependent. We assume density-dependent birth rate, ($B$), and death rate ($D$) in a discrete-time version of the classical Lotka--Volterra model [@de2013predicting; @vasseur2020impact] to get instantaneous growth rate, $\tilde{r}_{i}(t)$, for species, $i$, in year $t$:

```{=tex}
\begin{equation}\label{eq.r}
   \tilde{r}_{i}(t) = ln N_{i}(t+1) - ln N_{i}(t) = B(N_{i}(t),N_{j}(t),T(t)) - D(N_{i}(t),N_{j}(t),T(t))(\#eq:r)
\end{equation}
```
Here, $N(t)$ represents the biomass at year $t$, and $i,j$ are indices for two different species.

The per-capita birth and death rates for $i^{th}$ species are represented as:

```{=tex}
\begin{equation}\label{eq.B}
   B_{i} = b_{0,i}(T)-\beta (N_{i}+\sum_{i \neq j = 1}^{S} \alpha_{ij}N_{j})(\#eq:B)
\end{equation}
```
```{=tex}
\begin{equation}\label{eq.D}
   D_{i} = d_{0,i}(T)-\delta (N_{i}+\sum_{i \neq j = 1}^{S} \alpha_{ij}N_{j})(\#eq:D)
\end{equation}
```
where, $\beta$ and $\delta$ are density-dependent constants, $\alpha_{i,j}$ is the competition coefficient between species $i$ and $j$, and

```{=tex}
\begin{equation}\label{eq.b0}
   b_{0,i}(T) = a_{b} e^{-(T-b_{opt,i})^2/s_{i}} (\#eq:b0)
\end{equation}
```
```{=tex}
\begin{equation}\label{eq.d0}
   d_{0,i}(T) = a_{d} e^{z_{i}T} (\#eq:d0)
\end{equation}
```
with $a_{b}$, $a_{d}$ as intercepts, and for $i^{th}$ species, $b_{opt,i}$ is the temperature that optimizes birth rate, $s_{i}$ governs the breadth of the birth function, $z_{i}$ scales the effect of temperature (in °C) to mimic the Arrhenius relationship.

Substituting Eqs. (\ref{eq.B}) - (\ref{eq.d0}) in Eq. (\ref{eq.r}), we get the following,

```{=tex}
\begin{equation}\label{eq.r2}
   \tilde{r}_{i}(t) = r_{m,i} \left( 1-\dfrac{\sum_{i, j = 1}^{S} \alpha_{ij}N_{j}}{K_{i}} \right) (\#eq:r2)
\end{equation}
```
where, $\alpha_{ii} = 1$, $r_{m,i} = (b_{0,i} - d_{0,i})$ is the intrinsic (maximum) rate of increase and $K_{i} = r_{m,i}/(\beta+\delta)$ is the carrying capacity for the $i^{th}$ species, respectively.

At each time step there is a small amount of immigration into the system. This is implemented by adding 0.1 to all species' biomass at each time step. Done in the `simulator_lv` function.



## Model parameters

The following need to be specified for a community (in addition to the number of species):

```{r c4}
#| echo: false
#| eval: true
species_and_community_properties <- read.csv("species_and_community_properties.csv")
species_and_community_properties |> 
  kableExtra::kbl() |> 
  kableExtra::kable_styling()
```

# Designing an Experiment

An experiment consists of a number of "cases". Each case consists of a combination of community composition and environmental variability. The community composition is defined by the species present and their properties. The environmental variability is a time series of environmental conditions that the community is exposed to. When a single community has multiple replicates, then each replicate has the same community composition but different environmental variability.

The features of the experiment are given in a JSON file.



## Creating environmental variability

Currently created by 1 over f process with *gamma* (slope of frequency-power relationship) given in the json file. The 1 over f process is created with the `one_over_f` function from the (`primer` package)[https://search.r-project.org/CRAN/refmans/primer/html/00Index.html]. 

Also given in the json file are:

* The mean of the environmental time series.
* The standard deviation of the environmental time series.
* How the environmental time series vary among cases.

Currently three options for how environmental time series vary among cases.

1) `same_per_replicate` replicates of the same number share the same environmental time series. E.g., `community1_rep1` and `case2_rep1` share the same time series.
2) `all_different` all simulations have different environmental time series



# Typical Experiment Workflow

The typical workflow is as follows:

1. Set a project folder that can contain multiple experiment subfolders.
2. Set an experiment name and create a folder for that experiment.
3. Copy an example experiment-definition JSON file into the experiment folder.
4. Edit the JSON file to define the experiment you want to run.
5. Run the experiment workflow.
6. Inspect saved outputs and make plots.

## Project Folder and Experiment Folder

One way to organise work is to create a project folder that contains multiple
experiment subfolders:

```{r}
#| eval = FALSE
project_folder_location <- file.path("~/Desktop", "community_simulator_projects")
dir.create(project_folder_location, recursive = TRUE, showWarnings = FALSE)
```

## Copy an Example JSON Design File

The package includes bundled example experiment-definition files. A convenient
way to create an experiment folder and copy one of those JSON files into it is
to use `setup_example_experiment()`:

```{r}
#| eval = FALSE
project_folder_location <- file.path("~/Desktop", "community_simulator_projects")
experiment_name <- "discrete_lv_example"
experiment_design_filename <- "experiment_definition.json"

setup <- setup_example_experiment(
  experiment_folder_location = project_folder_location,
  experiment_name = experiment_name,
  example_experiment_name = "discrete_lv",
  experiment_design_filename = experiment_design_filename
)
```

This creates the experiment folder and copies the example JSON into it. At that
point, edit the JSON file in the experiment folder so it matches the experiment
you want to run.

The package includes three compact example designs, one for each supported
model type:

- `inst/test_experiments/discrete_lv/experiment_definition.json`
- `inst/test_experiments/continuous_lv/experiment_definition.json`
- `inst/test_experiments/consumer_resource/experiment_definition.json`

## Run the Experiment

Once the JSON file is in place and edited as needed, run the workflow with the
high-level wrapper:

```{r}
#| eval = FALSE
outputs <- run_experiment(
  experiment_folder_location = project_folder_location,
  experiment_name = experiment_name,
  experiment_design_filename = experiment_design_filename,
  overwrite = FALSE
)
```

By default, workflow functions stop if an output file already exists. To
replace existing outputs deliberately, re-run with `overwrite = TRUE`.

The main outputs are saved inside the experiment folder and typically include:

- `experiment_table.RDS`
- `temperatures.db`
- `dynamics.db`
- `community_measures.RDS`

# JSON File Specification

The design of an experiment is given in a JSON file.

An example of an experiment design file is included in the package. If you want
to copy it manually rather than using `setup_example_experiment()`, you can do
so like this:

```{r}
#| eval = FALSE
path_to_example_experiment_design_file <- system.file(file.path("test_experiments",
                      "discrete_lv",
                      "experiment_definition.json"),
            package = "community.simulator")
file_to_copy_to <- file.path("~/Desktop", "experiment_definition.json")
file.copy(from = path_to_example_experiment_design_file, to = file_to_copy_to)

```


### Allowed values

All values in that file are parsed and then must be R expressions that can be evaluated.

**Single numbers**

"name": 0

```{r}
eval(parse(text = "0"))
```

**Multiple numbers and functions**

"name": "c(0, 1, 2)"

```{r}
eval(parse(text = "c(0, 1, 2)"))
```

"name": "0:10"

```{r}
eval(parse(text = "0:10"))
```

"name": "seq(0, 10, 2)"

```{r}
eval(parse(text = "seq(0, 10, 2)"))
```



**Single words**

Do not use: "name": "word"

```{r eval = FALSE}
eval(parse(text = "word"))
```

It must be:

"name": "\"word\""

```{r}
eval(parse(text = "\"word\""))
```

**Multiple words**

"name": "c(\"word1\", \"word2\")"

```{r}
eval(parse(text = "c(\"word1\", \"word2\")"))
```






## Installation

Install from GitHub:

```r
install.packages("remotes")
remotes::install_github("opetchey/community.simulator")
```

Install from GitHub and build the vignettes during installation:

```r
install.packages(c("remotes", "knitr", "rmarkdown"))
remotes::install_github(
  "opetchey/community.simulator",
  build_vignettes = TRUE,
  dependencies = TRUE,
  upgrade = "never"
)
```

Install from a local checkout:

```r
install.packages("remotes")
remotes::install_local(".")
```

Install from a local checkout and build the vignettes during installation:

```r
install.packages(c("remotes", "knitr", "rmarkdown"))
remotes::install_local(
  ".",
  build_vignettes = TRUE,
  dependencies = TRUE,
  upgrade = "never"
)
```

After installation, you can view the installed vignettes with:

```r
browseVignettes("community.simulator")
```

You can also open the current user guide directly:

```r
vignette("experiment_getting_started", package = "community.simulator")
vignette("User_guide", package = "community.simulator")
```

If you are working interactively inside the repository during development:

```r
devtools::load_all()
```

## Quick Example

The typical workflow is:

1. Set a project folder that can contain multiple experiment subfolders.
2. Set an experiment name and create a folder for that experiment.
3. Copy an example experiment-definition JSON file into the experiment folder.
4. Edit the JSON file to define the experiment you want to run.
5. Run the experiment workflow.
6. Inspect plots and saved outputs.

The main user-facing functions are:

- `run_experiment()`
- `setup_example_experiment()`
- `create_experiment_folder()`
- `create_experiment_table()`
- `create_environments()`
- `simulate_dynamics()`
- `get_community_measures()`
- `make_plots_for_one_community()`

Bundled example experiment designs are available under `inst/test_experiments`.
The package currently ships one compact design for each supported model type:

- `discrete_lv/experiment_definition.json`
- `continuous_lv/experiment_definition.json`
- `consumer_resource/experiment_definition.json`

One way to get started is to create a project folder on your Desktop that can
contain multiple experiments:

```r
project_folder_location <- file.path("~/Desktop", "community_simulator_projects")
dir.create(project_folder_location, recursive = TRUE, showWarnings = FALSE)
```

Then create a folder for a single experiment:

```r
library(community.simulator)

project_folder_location <- file.path("~/Desktop", "community_simulator_projects")
experiment_name <- "discrete_lv_example"
experiment_design_filename <- "experiment_definition.json"

setup <- setup_example_experiment(
  experiment_folder_location = project_folder_location,
  experiment_name = experiment_name,
  example_experiment_name = "discrete_lv",
  experiment_design_filename = experiment_design_filename
)
```

At this point, edit the JSON file in the experiment folder so it matches the
experiment you want to run.

Then run the experiment:

```r
outputs <- run_experiment(
  experiment_folder_location = project_folder_location,
  experiment_name = experiment_name,
  experiment_design_filename = experiment_design_filename,
  overwrite = FALSE
)
```

By default, workflow functions now stop if an output file already exists. To
replace existing outputs deliberately, re-run with `overwrite = TRUE`.

For a more guided version of this example, see
[vignettes/experiment_getting_started.Rmd](vignettes/experiment_getting_started.Rmd).

## Data and Outputs

The workflow saves intermediate and final outputs inside the experiment folder.
These currently include:

- `experiment_table.RDS`
- `experiment_log.txt`
- `temperatures.db`
- `dynamics.db`
- `community_measures.RDS`

Some optional downstream steps also create additional derivative or diagnostic
databases.

## Learning One Model At A Time

The QMD files in [reports/examples](reports/examples) define one community,
one temperature environment, and one simulation directly in R. They are intended
for learning and experimentation rather than large simulation studies.

- [Discrete-time LV walkthrough](reports/examples/lv_discrete_single_community.qmd)
- [Continuous-time LV walkthrough](reports/examples/lv_continuous_single_community.qmd)
- [Consumer-resource walkthrough](reports/examples/consumer_resource_single_community.qmd)

Render any of them from the package root with:

```r
quarto::quarto_render("reports/examples/lv_discrete_single_community.qmd")
```

