---
title: "YAML Experiment Templates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{YAML Experiment Templates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Purpose

The package includes YAML templates for the three supported model families.
These templates are the recommended starting point for experiment-level work:
copy one, edit the values, validate it with `read_experiment_spec()`, and run it
with `run_experiment()`.

The compact templates are deliberately small. They are useful for smoke tests,
reviewer checks, and learning the shape of the specification.

| Model | Compact template |
| --- | --- |
| Discrete-time LV | [`lv_discrete.yaml`](https://github.com/opetchey/community.simulator/blob/main/inst/experiment_templates/lv_discrete.yaml) |
| Continuous-time LV | [`lv_continuous.yaml`](https://github.com/opetchey/community.simulator/blob/main/inst/experiment_templates/lv_continuous.yaml) |
| Continuous-time consumer-resource | [`consumer_resource.yaml`](https://github.com/opetchey/community.simulator/blob/main/inst/experiment_templates/consumer_resource.yaml) |

The rich templates show fuller experiment specifications, including treatments,
replicates, interaction or resource-use options, output controls, and parallel
settings.

| Model | Rich template |
| --- | --- |
| Discrete-time LV | [`lv_discrete_rich.yaml`](https://github.com/opetchey/community.simulator/blob/main/inst/experiment_templates/lv_discrete_rich.yaml) |
| Continuous-time LV | [`lv_continuous_rich.yaml`](https://github.com/opetchey/community.simulator/blob/main/inst/experiment_templates/lv_continuous_rich.yaml) |
| Continuous-time consumer-resource | [`consumer_resource_rich.yaml`](https://github.com/opetchey/community.simulator/blob/main/inst/experiment_templates/consumer_resource_rich.yaml) |

## Copy A Template

After installing the package, locate a bundled template with `system.file()`:

```{r, eval = FALSE}
template <- system.file(
  "experiment_templates/lv_discrete_rich.yaml",
  package = "community.simulator"
)
```

For a ready-made experiment folder, use `setup_example_experiment()`:

```{r, eval = FALSE}
setup_example_experiment(
  experiment_folder_location = tempdir(),
  experiment_name = "lv_discrete_review",
  example_experiment_name = "lv_discrete_rich"
)
```

Then run the copied YAML specification with `run_experiment()`.

