---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette runs a very small end-to-end experiment using
`community.simulator`. The goal is to show the simplest successful workflow for
new users.

The example:

- copies a bundled experiment-definition file to a temporary experiment folder
- runs the standard workflow with `run_experiment()`
- inspects the saved summary output

```{r setup}
library(community.simulator)
```

## Run a Tiny Example

```{r}
experiment_folder_location <- tempdir()
experiment_name <- "getting_started_example"
experiment_design_filename <- "getting_started_experiment.json"

experiment_folder <- create_experiment_folder(
  experiment_folder_location = experiment_folder_location,
  experiment_name = experiment_name,
  verbose = FALSE
)

design_source <- system.file(
  "extdata",
  experiment_design_filename,
  package = "community.simulator"
)

file.copy(
  from = design_source,
  to = file.path(experiment_folder, experiment_design_filename),
  overwrite = TRUE
)

outputs <- suppressWarnings(
  suppressMessages(
    run_experiment(
      experiment_folder_location = experiment_folder_location,
      experiment_name = experiment_name,
      experiment_design_filename = experiment_design_filename,
      overwrite = TRUE,
      verbose = FALSE
    )
  )
)

outputs
```

## Inspect the Main Output

```{r}
community_measures <- readRDS(outputs$community_measures)

community_measures[, c("case_id", "richness", "temperature_mean", "CV_totab")]
```

## Next Steps

Once this small example works, the usual next steps are:

- edit an experiment-definition file to change species, temperature, or
  replicate settings
- rerun the workflow in a new experiment folder
- inspect `community_measures.RDS`, `dynamics.db`, and `temperatures.db`
- use `make_plots_for_one_community()` for case-level plots
